\documentclass[addpoints,12pt]{exam}
\usepackage{color}
\usepackage{amsmath}

\begin{document}
\title{Quiz 1 Practice Questions}
\author{}
\date{}
\maketitle

\newcommand{\hint}[1]{\\{\bf Hint:} {\em {#1}}}



\fullwidth{\Large \textbf{Multiple Choice Questions}}


\begin{questions}
\qformat{[M\thequestion]}
\question
What is the recurrence relation for Merge Sort?
\begin{choices}
\choice $T(n) = T(n-1) + 1$
\choice $T(n) = T(n-1) + n$
\choice $T(n) = T(n/2) + n$
\choice $T(n) = 2T(n/2) + 1$
\choice None of the above 
\end{choices}



\question
What is the solution to the recurrence $T(n) = T(n-1) + 1$
\begin{choices}
\choice $O(n)$
\choice $O(n^2)$
\choice $O(n^3)$
\choice $O(2^n)$
\choice None of the above 
\end{choices}



\question
What is the lower bound for sorting algorithms based on comparison?
\begin{choices}
\choice $\Omega(n \log n)$
\choice $\Theta(n \log n)$
\choice $\Omega(n^2)$
\choice $\Theta(n^2)$
\choice None of the above 
\end{choices}



\question
Suppose you always pick median as the pivot in the QuickSort. What is the recurrence relation for this variant?
\begin{choices}
\choice $T(n) = T(n-1) + 1$
\choice $T(n) = T(n-1) + n$
\choice $T(n) = T(n/2) + n$
\choice $T(n) = 2T(n/2) + 1$
\choice None of the above 
\end{choices}



\question
What is the time complexity of Radix sort? Suppose that the array has $n$ elements with $d$ digits each. Also each digit can take values between $1$ and $k$
\begin{choices}
\choice $O(n+k)$
\choice $O(n+k+d)$ 
\choice $O(n(k+d))$
\choice $O(d(n+k))$
\choice None of the above 
\end{choices}
\end{questions}




\begin{center}
{\Large \textbf{Algorithmic Questions}}
\end{center}


\fullwidth{\Large \textbf{Topic : General}}
\begin{questions}
\qformat{[G\thequestion]}
\question
You are given 9 identical looking balls and are told that one of them weighs a bit less than the rest of 
the eight balls. The only operation you are allowed is to compare a set of balls against another set of 
balls. Determine the lighter ball using 3 comparisons. Generalize your answer to more that 9 balls if 
possible.

\question
You are given 12 balls, and are told that one of them is of a different weight from the rest - 
i.e., you don't know if it is heavier or lighter. Determine this ball using 4 comparisons.
\end{questions}


\fullwidth{\Large \textbf{Topic : Asymptotics}}

\begin{questions}
\qformat{[A\thequestion]}

\question
Solve the following recurrence relations:
\begin{choices}
\choice $T(n) = 2T(n/3) + n^2$ 
\choice $T(n) = 2T(n-1)$
\choice $T(n) = T(n/2) + 1$
\choice $T(n) = 4T(n/2) + n$
\end{choices}

\question
 Compare the following functions in terms of orders. In each case, say whether 
$f(n) = O(g(n))$, $f(n) = \Theta(f(n))$ or $f(n) = \Omega(f(n))$
\begin{choices}
    \choice $f(n) = \sqrt{n}$ and $g(n) = (\log n)^5$
    \choice $f(n) = n^2 \log n $ and $g(n) = n (\log n)^2$
    \choice $f(n) = \log n$ and $g(n) = (\log n)^2$
    \choice $f(n) = n \, 2^n$ and $g(n) = 3^n$
\end{choices}

\end{questions}

\fullwidth{\Large \textbf{Topic : Order Statistics}}
\begin{questions}
\qformat{[OS\thequestion]}
\question
The majority of a set of numbers is defined as a number that repeats at least $\frac{n}{2}$ times in the set. 
Design a linear time algorithm to find the majority, {\bf if} one exists.

\question
Show exactly why if we grouped elements into groups of 3 each, the median finding algorithm that 
we discussed in class will not work in linear time. What would be the running time of the algorithm in 
this case?

\question
Which is better in the median-finding algorithm, grouping into groups of 5, or into groups of 7? 
Explain your answer.

\question
Let $X$ and $Y$ be two arrays of $n$ numbers each, both already sorted. Give a $O(\log n)$ algorithm to 
compute the median of $X \cup Y$.


\end{questions}



\fullwidth{\Large \textbf{Topic : Sorting}}
Which sorting algorithm would you use for the following cases?
\begin{questions}
\qformat{[S\thequestion]}

\question
The array is almost sorted.

\question
The array is sorted in reverse order.

\question
Sorting UTA student ids (1000nnnnnn).

\question
Sorting mp3 files in your player based on the name. 
\hint{Note: This is a case where you want to minimize number of swaps - each file is, say 5MB. Then doing unnecessary swapping
will degrade the SSD of your drive.}

\question
Your array is too huge to hold in memory.

\question
You are given an array that was randomly shuffled. 

\question
The size of your array is close to the size of your RAM.

\question
Sorting cheques in a bank based on the cheque id.
\hint{Note that most customers use cheques in the order given in the cheque book.}

\end{questions}


\fullwidth{\Large \textbf{Topic : Trees}}
\begin{questions}
\qformat{[G\thequestion]}
\question
Suppose you are given a completely balanced binary search tree (a completely balanced tree 
means that each path from root to leaf is exactly the same). What is the time required to find the median
of the elements of such a tree ?

\question
Suppose you are given a completely balanced binary search tree (a completely balanced tree 
means that each path from root to leaf is exactly the same). What is the time required to find the max 
of the elements of such a tree ?

\question
Suppose some one gave you a binary tree and claimed that it is a BST. 
Design and analyze an efficient algorithm to verify if it is indeed the case.

\question
Given a BST $T$ and two integers $a$ and $b$, print all elements in $T$ between $a$ and $b$.

\question
Suppose you are given two trees $T_1$ and $T_2$. Design an efficient algorithm to determine if they contain the same elements.

\question
Design an algorithm to convert a {\em sorted} array to a {\em balanced} binary tree.

\end{questions}

\fullwidth{\Large \textbf{Topic : Potpourri}}
\begin{questions}
\qformat{[P\thequestion]}

\question
The input is a set $S$ containing $n$ numbers, and a number $x$
\begin{parts}
\part
Design an algorithm to determine whether there are two elements of $S$ whose sum is exactly $x$. 
The algorithm should run in $O(n \log n)$ time.
\part
Suppose now that the set $S$ is given in a sorted order. Design an algorithm to solve the above problem in time $O(n)$.
\end{parts}

\question
Suppose you are given a sorted array $A$. Design an efficient algorithm to find if an index $i$ exists such that $A[i]=i$.
\end{questions}



\end{document}

